The Kapowski Chronicles: A Complete, Volumetric Cortical Thickness Open Source Pipeline with Evaluation on Public Data

### 2013-05-08 ANTsRS/paper.tex
ENH:  Added figure.
ENH:  Amy

\title{Multivariate Neuroanalysis with ANTsR:  Application to Supervised Brain Segmentation}

@@

O

@@

@@

Building on the developments of Ho \cite{ho1995} and Amit and Geman \cite{amit1997},
Breiman introduced Random Forests for classification/regression \cite{breiman2001} which performs relatively well against other machine learning strategies such as support vector machines and neural networks.

\subsubsection{Multi-Modality Feature Image Preprocessing}

Although several studies have pointed out the importance of
intensity normalization and bias correction, our experience 
with the training data illustrated a degradation in performance
when one or both steps (using \cite{nyul2000} and \cite{tustison2010},
respectively) were performed due to the presence of the tumor/edema complex.  
Instead we simply windowed the image intensity
for all images to be between the quantiles $[0.01,0.99]$ and
subsequently rescaled to $[0,1]$.  This strategy seemed to provide 
the best classification (probably due to the fact that the intensity
corruption due to bias was relatively minimal in the data used).
However, improved results might be obtained
by a more sophisticated strategy in which an initial supervised segmentation 
is used to find the probabilistic estimates of the normal tissue
locations (as done in this work) which could then be followed by using
only the probabilistic maps of the normal tissues to construct a
weight, or confidence, mask for estimating the inhomogeneity bias
using N4 \cite{tustison2010} followed by a repeated application of 
the supervised segmentation strategy described.  Note that a similar
iterative-based strategy until convergence is implemented in the ANTs script \verb#antsAtroposN4.sh#.

\subsubsection{Multi-Modality Feature Image Generation}

@@

\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius = 3.
    \item GMM-based posteriors: CSF, gray matter, white matter, edema, and tumor
    \item GMM connected component geometry features:  volume, volume to 
          surface area ratio, eccentricity, elongation
    \item Template-based:  symmetric template difference and contralateral difference.
          Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Misc.
    \begin{itemize}
    \item Normalized Euclidean distance
    \item Log Jacobian  
    \end{itemize}
\end{itemize}

For each modality, we create four first-order statistical feature images,
five Gaussian mixture model (GMM)-based posterior probability feature images,
four geometry features generated from the GMM posterior probability images
based on connected components, and two difference images using symmetric template
construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 images.  We employ two additional images consisting of the 
Euclidean distance image \cite{maurer2003} created from the skull-stripped 
binary mask rescaled to the range $[0,1]$ and the log
Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
the corresponding multivariate template images, and a brain mask for each subject
creation of all feature images is performed using the script \verb#createFeatureImages.sh#.

@@

\begin{table*}
\caption{Dice scores from the MICCAI 2012 BRATs Study}
\begin{center}
\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
\toprule
{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
\midrule
\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
\rowcolor{yellow}
ANTsR & {$0.64 \pm 0.18$} & {$0.65 \pm 0.30$} & {$0.48 \pm 0.16$} & {$0.69 \pm 0.21$} & {$0.75 \pm 0.25$} & {$0.94 \pm 0.04$} & {$0.65 \pm 0.25$} & {$0.83 \pm 0.09$}  \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

>>>>

\title{Multivariate Neuroanalysis with ANTsR:  Application to Supervised Brain Segmentation with Concatenated Random Forests}

@@

ANTs (Advanced Normalization Tools) originated with the open source availing
of state-of-the-art registration algorithms for neuroimage analysis
\cite{avants2008a} built upon the mature and well-vetted Insight Toolkit
of the National Institutes of Health.  Since then, the toolkit has grown to include 
several algorithmic solutions necessary
for robust medical image analysis including bias correction \cite{tustison2010}, 
$n$-tissue multivariate segmentation \cite{avants2011}, template construction \cite{avants2010}, 
and cortical thickness estimation \cite{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law).%
\footnote{
``Given enough eyeballs, all bugs are shallow.'' --Linus Torvalds
}   
However, in the evolution of the toolkit, it became clear (as neuroimaging
research certainly falls under the popular rubric of ``big data analysis'')
that robust statistical machinery was lacking for proper inferences regarding
the data produced by the various ANTs tools.
ANTsR was developed
specifically to provide an interface between a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations and the R project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate image analysis.

In addition to describing ANTsR basics and how it can be generally
used in multivariate neuroimaging studies, we showcase its use in
a particularly salient application for performing supervised brain
segmentation using random forests.  Although random forests have
been proposed previously in the literature for supervised brain 
segmentation (e.g. \cite{geremia2011,zikic2012}), we demonstrate that ANTsR
significantly facilitates the construction of a fully functional, 
parallelizable workflow for such a task and that performance 
exceeds that of the top competitors in the recent Multimodal Brain
Tumor Segmentation Challenge as part of the MICCAI 2012 conference.%
\footnote{
http://www2.imm.dtu.dk/projects/BRATS2012/  
}

@@

Given a set of training data consisting of 
labeled (csf, gray matter, white matter, edema, and tumor),
multimodal brain image data, 
supervised brain segmentation entails the generation of a 
statistical model from such training data which can
then be extended to testing data, i.e. unlabeled 
brain data.  In subsequent sections, we introduce the generic
{\it random forest} modeling framework which takes as input
a set of {\it feature images} (also described in a later section)
and labels for each training subject 
and outputs a statistical model.  This model, in turn, can then
be used to provide a probabilistic estimate of the labels in an
unlabeled subject.

One of the core extensions that we provide in this work is
the use of concatenated random forests for improved probabilistic estimation
of the labels.
As we demonstrate in the Results section, the set of feature
images employed work sufficiently well for good performance
on the training data (which exceeds that of what has been
currently published in the literature).  However, we discovered
that further improvements could be gained by using the probabilistic 
label estimates as input to enhance preprocessing before a 
second round of feature image generation including 
modality-specific, prior-based $n$-tissue segmentation and 
random forest model creation.

@@

Several machine learning concepts were integrated to create 
the random forests framework first articulated in its entirety by Breiman
et al. \cite{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \cite{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \cite{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \cite{breiman1996} to achieve improved
accuracy.

Early adoption \cite{viola2005} and success in the
computer vision community
has led to a recent surge within the medical image analysis
community of using random forests for handling complex 
classification/regression tasks including
normal brain segmentation \cite{yi2009},
MS lesion segmentation \cite{geremia2011}, 
multimodal brain tumor segmentation
\cite{zikic2012,geremia2012}, brain extraction \cite{iglesias2010}, 
anatomy detection in computed tomography \cite{criminisi2013}, and
segmentation of echocardiographic images \cite{verhoek2011}.
A thorough introduction for those interested in delving deeper 
into the more theoretical aspects of random forests can be found
in \cite{criminisi2011}.

One of the principal advantages of R is the extensive community of
developers  who have contributed on the order of thousands of packages 
extending R's capabilities beyond its core functionality.
Most relevant for the work described in this paper
is the \verb#randomForest# package developed from Breiman's original
Fortran code by Liaw and Wiener (described briefly in \cite{liaw2002}).

\subsubsection{Multi-Modal Feature Image Preprocessing}

Although several studies have pointed out the importance of
intensity normalization and bias correction, our experience 
with the training data illustrated a degradation in performance
when one or both steps (using \cite{nyul2000} and N4 \cite{tustison2010},
respectively) were performed due to the presence of the tumor/edema complex.

As a corrective, for the first stage we simply windowed the image intensity
for all images to be between the quantiles $[0.01,0.99]$ and
subsequently rescaled to $[0,1]$.  From these ``corrected'' images,
the first set of feature images were derived.  For the training cohort, these
data were used to create the random forest regression model for the first
stage.  During the second stage, the probabilistic estimates
of the white matter and gray matter labels were used to generate a
``pure tissue weight mask'' to estimate the bias field 
using N4 (although the resulting bias field estimation was used
to correct the image within the entire cerebral mask).  Formally, this 
involved generation of a probabilistic map defined as:
\begin{align}
  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right)
\end{align}
where $N$ is the set of user-selected tissue labels (in our
case $N=2$ consisting of the gray and white matter probability
maps).

Both rescaling and weighted bias correction were applied to produce
the ``corrected images'' for the second stage resulting in
modified features images for the second stage.  Note that we
perform a similar iterative scenario for normal brain 
segmentation \cite{avants2011} (encapsulated in the ANTs script 
\verb#antsAtroposN4.sh#).

\subsubsection{Multi-Modal Feature Image Generation}

@@

\begin{figure*}
  \centering
  \includegraphics[width=180mm]{Figures/featuresImages.pdf}
  \caption{Feature images from the BRATS\_HG0004 data set.}
\end{figure*}

\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius = 3.
    \item GMM-based posteriors: CSF, gray matter, white matter, edema, and tumor
    \item GMM connected component geometry features:  volume, volume to 
          surface area ratio, eccentricity, elongation
    \item Template-based:  symmetric template difference and contralateral difference.
          Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Not modality specific
    \begin{itemize}
    \item Normalized Euclidean distance
    \item Log Jacobian  
    \item T1C,T1 difference image
    \end{itemize}
\end{itemize}

For each modality, we create four first-order statistical feature images,
five Gaussian mixture model (GMM)-based posterior probability feature images,
four geometry features generated from the GMM posterior probability images
based on connected components, and two difference images using symmetric template
construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 total
feature images.  We employ two additional images consisting of the 
Euclidean distance image \cite{maurer2003} created from the skull-stripped 
binary mask rescaled to the range $[0,1]$ and the log
Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
the corresponding multivariate template images, and a brain mask for each subject
creation of all feature images is performed using the script \verb#createFeatureImages.sh#.

@@

\begin{table*}
\caption{Dice scores from the MICCAI 2012 BRATs Study}
\begin{center}
\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
\toprule
{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
\midrule
\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
ANTsR & {$0.65 \pm 0.15$} & {$0.66 \pm 0.28$} & {$0.49 \pm 0.16$} & {$0.65 \pm 0.21$} & {$0.68 \pm 0.25$} & {$0.91 \pm 0.08$} & {$0.61 \pm 0.25$} & {$0.84 \pm 0.09$} \\
w/ Atropos & {$0.68 \pm 0.15$} & {$0.67 \pm 0.30$} & {$0.50 \pm 0.15$} & {$0.67 \pm 0.23$} & {$0.74 \pm 0.26$} & {$0.92 \pm 0.09$} & {$0.65 \pm 0.26$} & {$0.84\pm 0.08$} \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

### 2013-05-08 KapowskiChronicles/intro.tex
ENH:  Amy

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures are able to identify abnormalities in such conditions as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011}, Williams syndrome \cite{thompson2005},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, untreated transsexuality \citep{zubiaurre-elorza2012},  handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

Image volumetric (or meshless) techniques vary both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation where thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are potentially problematic as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms employ coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed by \cite{das2009} in which 
a registration-based approach proposed where the derived diffeomorphic mapping between the 
white and pial matter surfaces is used to propagate thickness values 
through the cortical gray matter.  A unique benefit of DiReCT is that it
naturally estimates the boundaries of buried sulci by employing a
diffeomorphic constraint on the probabilistic estimate of the gray
matter and cerebrospinal fluid interface.

Despite the variety of techniques for estimating cortical thickness
from imaging data (of which
only a fraction are cited), several common pre-processing components
may be identified.
The most fundamental of these include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matter.  For statistical analysis 
across large populations, construction of population-specific unbiased templates
is also beneficial \cite{evans2012}.
In addition, some of these steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.  The requisite
additional components coupled with the general lack of availability of published
algorithms \cite{kovacevic2006} inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual algorithmic developers) using 
unspecified parameters making the comparisons less than
ideal. Further complicating comparisons is distinct processing domains between
volumetric and surface-based techniques and the potential for the introduction
of bias \cite{klein2010}.

>>>>

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures have been demonstrated in conditional 
abnormalities such as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011}, Williams syndrome \cite{thompson2005},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, untreated transsexuality \citep{zubiaurre-elorza2012},  handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

Image volumetric (or meshless) techniques vary both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation where thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are potentially problematic as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms employ coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed in \cite{das2009} where generated
diffeomorphic mappings between the 
white and pial matter surfaces are used to propagate thickness values 
through the cortical gray matter.  A unique benefit of DiReCT is that it
naturally estimates the boundaries of buried sulci by employing a
diffeomorphic constraint on the probabilistic estimate of the gray
matter and cerebrospinal fluid interface.

Despite the variety of techniques for estimating cortical thickness
from imaging data (of which
only a fraction are cited), several common preprocessing components
may be identified.
The most fundamental of these include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matter.  For statistical analysis 
across large populations, construction of population-specific unbiased templates
is also potentially beneficial \cite{evans2012}.
In addition, intermediate steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.

  The requisite
additional components apart from the actual cortical thickness estimation, 
coupled with the general lack of availability of published
algorithms \cite{kovacevic2006}, inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual algorithmic developers) using 
unspecified parameters making the comparisons less than
ideal. Further complicating comparisons is distinct processing domains between
volumetric and surface-based techniques and the potential for the introduction
of bias \cite{klein2010}.
